Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-04-24 marginal.Rmd
Some final corrections and additions.

Social position affects health and mortality differently during the life course. There are different hypotheses about the impact in adulthood and old age respectively. The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes 
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral rewarding some while disfavouring others. This leads to larger differences in old age.

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. The regions studied here were for a long time remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships of the year came to this area in May or even as late as June in some cases [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden and, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

@@

The other large data set is extracted from the Linnaeus database
[@malm10], which is based on different linked national population
registers from 1960 to 2013 (censuses, LISA from Statistics Sweden and cause of death registers) and is used within the ageing programme at CEDAR, Umeå
University. Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data from 1976 onwards. The study period from 1976 to 2000 is constructed from
censuses every fifth year during the period 1975--1990, with additional information on deaths
from the National Board of Health and Welfare. For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with information on mortality from the National Board of Health and
Welfare.

@@

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

Figure \@ref(fig:probswabs) shows the absolute risks of dying for in the age interval 40--64 during the different periods. Obviously, but not surprisingly, mortality
decreases over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history [@wil99; @suwi07] as well as in the rest of the world. The
substantial decline in mortality for both sexes results in smaller
absolute differences between social classes in the later time periods. We must, however, distinguish between differences in absolute risks and in relative risks. The compression that is visible for later decades is partly explained by the decreasing general mortality and does not necessarily represent a homogenisation of relative risks [@hoetal16]. The relative risks are shown in Figure \@ref(fig:probsw).

@@

>>>>

Social position affects health and mortality differently during the life course. There are different hypotheses about the impact in adulthood and old age respectively. The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes 
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral, rewarding some while disfavouring others. This leads to larger differences in old age.

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. The regions studied here were for a long time remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships of the year came to this area in May or even as late as June in some cases [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

@@

The other large data set is extracted from the Linnaeus database
[@malm10], which is based on different linked national population
registers from 1960 to 2013 (censuses, LISA from Statistics Sweden and cause of death registers) and is used within the ageing programme at CEDAR, Umeå
University. Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data only from 1976 onwards. The study period from 1976 to 2000 is constructed from
censuses every fifth year during the period 1975--1990, with additional information on deaths
from the National Board of Health and Welfare. For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with information on mortality from the National Board of Health and
Welfare.

@@

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

Figure \@ref(fig:probswabs) shows the absolute risks of dying for in the age interval 40--64 during the different periods. Obviously, but not surprisingly, mortality
decreases over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history [@wil99; @suwi07] as well as in other parts of the world. The
substantial decline in mortality for both sexes results in smaller
absolute differences between social classes in the later time periods. We must, however, distinguish between differences in absolute risks and in relative risks. The compression that is visible for later decades is partly explained by the decreasing general mortality and does not necessarily represent a homogenisation of relative risks [@hoetal16]. The relative risks are shown in Figure \@ref(fig:probsw).

@@

Earlier versions have been presented at 11th European Social Science History Conference, Valencia 2016, 2nd Conference of the European Society of Historical Demography, Leuven 2016, 41st Annual Meeting of the Social Science History Association, Chicago 2016, The Ninth International Congress of Arctic Social Sciences ICASS IX, Umeå 2017, 20th Nordic Demographic Symposium, Åbo 2017 and XXVII International Population Conference, Cape Town 2017. We are grateful for comments and suggestions at these conferences as well as comments from our colleagues ag CEDAR, Umeå University.
